Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-05 ms.qmd
Some more writing

Arguably, reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are likely to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Change in physiological variability can also inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance depending on the trait [@Scheiner2020]. The implications of changes in variance could therefore have wide-reaching implications for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., see @Seebacher2015-wi], it is unclear if the magnitude of physiological adjustment is sufficient. In addition, low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits is expected to be a stronger constraint on physiological processes in terrestrial ectotherms.

>>>>

Arguably, reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are likely to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010; @Zhang2023]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020]. The implications of changes in variance could therefore have wide-reaching implications for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., see @Seebacher2015-wi], it is unclear if the magnitude of physiological adjustment is sufficient. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits is expected to be a stronger constraint on physiological processes in terrestrial ectotherms. Given that terrestrial ectotherms are expected to be closer to their thermal limits increases in temperature may have a stronger impact on variation in physiological rates within populations compared to aqautic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both.
